---
title: "*Bradyrhizobium* diversity: curation, phylogeny and ecological distribution"
author: "Italo Dutra"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)

library(tidyverse)
library(janitor)
library(countrycode)
library(readxl)
library(writexl)
```

# Introduction and objective

This document records, in a reproducible way, the process of obtaining,
curating and analyzing *Bradyrhizobium* genomes deposited in NCBI, aiming to
assess:

1. geographic distribution (by continent) of genomes;
2. distribution by plant host genus;
3. distribution by taxonomic group within the genus (*B. japonicum*,
   *B. elkanii*, *B. diazoefficiens*).

The general methodology follows the logic described in Sobol et al. (2026,
*ISME Communications*, ycag007), with adaptations described throughout the
text — mainly in how genomes are selected (a dedicated catalog built via NCBI
Datasets CLI, instead of subsampling a pre-existing reference phylogeny) and
in how the phylogeny is built (from core genes, instead of functional
annotation over an already-established tree).

**Naming convention:** fields produced directly by NCBI's `dataformat` output
are kept as-is (already in English, e.g. `assembly_accession`,
`check_m_completeness`). Fields derived by us during curation (country,
continent, host genus, taxonomic group, etc.) are named in English from this
version of the document onward, to facilitate sharing this repository.

Bash code chunks document commands executed in the terminal (WSL/Ubuntu) and
are marked `eval=FALSE` (shown for documentation, not re-executed when
rendering, to avoid triggering large downloads every time). R chunks are
executed for real when rendering, reproducing the tables and numbers shown.

---

# Step 1 — Obtaining the genome list (NCBI)

```{bash, eval=FALSE}
conda create -n bradyrhizobium -c conda-forge -c bioconda \
  ncbi-datasets-cli entrez-direct -y
conda activate bradyrhizobium
```

```{bash, eval=FALSE}
datasets summary genome taxon "Bradyrhizobium" --as-json-lines \
  > bradyrhizobium_genomes.jsonl

wc -l bradyrhizobium_genomes.jsonl
# result: 4047 raw records
```

```{bash, eval=FALSE}
dataformat tsv genome --inputfile bradyrhizobium_genomes.jsonl \
  --fields accession,assminfo-level,assminfo-atypicalis-atypical,\
source_database,checkm-completeness,checkm-contamination,\
assminfo-biosample-host,assminfo-biosample-geo-loc-name,\
assminfo-biosample-isolation-source,assminfo-biosample-lat-lon,\
organism-name \
  > bradyrhizobium_diagnostico.tsv
```

---

# Step 2 — Reading, deduplication and quality filter

```{r}
genomes_raw <- read_tsv("bradyrhizobium_diagnostico.tsv", show_col_types = FALSE) |>
  clean_names()

names(genomes_raw)
```

## Removing GenBank/RefSeq duplicates

NCBI frequently deposits the same physical genome twice (once under GenBank,
prefix `GCA_`, and once under RefSeq, prefix `GCF_`, sharing the same base
accession number). We keep a single entry per physical genome, prioritizing
the RefSeq version when available.

```{r}
genomes_raw <- genomes_raw |>
  mutate(
    accession_number = str_remove(assembly_accession, "^GC[AF]_"),
    db_prefix        = str_sub(assembly_accession, 1, 3)
  )

genomes_dedup <- genomes_raw |>
  group_by(accession_number) |>
  arrange(desc(db_prefix == "GCF"), .by_group = TRUE) |>
  slice(1) |>
  ungroup()

nrow(genomes_raw)
nrow(genomes_dedup)
```

## Genomic quality filter

Criterion: completeness >=90% and contamination <5% (Bowers et al. 2017;
same criterion used in Sobol et al. 2026).

```{r}
genomes_curated <- genomes_dedup |>
  filter(check_m_completeness >= 90, check_m_contamination < 5)

nrow(genomes_curated)

genomes_curated |> count(assembly_level, sort = TRUE)
```

---

# Step 3 — Geographic standardization (country -> continent)

```{r}
genomes_curated <- genomes_curated |>
  mutate(
    country = str_trim(str_extract(assembly_bio_sample_geographic_location, "^[^:]+")),
    country_clean = case_when(
      country %in% c("missing", "Missing", "not applicable", "not provided",
                      "not recorded", "-", "None") ~ NA_character_,
      TRUE ~ country
    ),
    continent = countrycode(country_clean, origin = "country.name", destination = "continent"),
    continent = case_when(
      country_clean %in% c("Arctic Ocean", "Antarctica") ~ "Not applicable (environmental source)",
      TRUE ~ continent
    )
  )

genomes_curated |> count(continent, sort = TRUE)
```

**Decision:** the initial geographic analysis is conducted at the broad
continent level (Americas, Africa, Asia, Europe, Oceania). Sub-regional cuts
(e.g. South America, Africa specifically) or a "Global South" grouping — for
which no single consensus definition exists in the literature — are deferred
to a separate, later analysis with an explicitly stated criterion.

---

# Step 4 — Host standardization

```{r}
hosts_unique <- genomes_curated |> count(assembly_bio_sample_host, sort = TRUE)
```

```{r}
genomes_curated <- genomes_curated |>
  mutate(
    host_clean = assembly_bio_sample_host,
    host_clean = str_remove(host_clean, "\\(L\\.\\)\\s*Merr(ill)?\\.?"),
    host_clean = str_remove(host_clean, "\\(Sw\\.\\)\\s*Willd\\.?"),
    host_clean = str_remove(host_clean, "\\bcv\\.\\s*[A-Za-z ]+"),
    host_clean = str_remove(host_clean, "\\bcultivar\\s+[A-Za-z0-9]+"),
    host_clean = str_remove(host_clean, "\\bsubsp\\.\\s*[A-Za-z]+"),
    host_clean = str_remove(host_clean, "\\bvar\\.\\s*[A-Za-z]+"),
    host_clean = str_remove(host_clean, "\\bBenth\\.?"),
    host_clean = str_remove(host_clean, "\\bCol-0\\b"),
    host_clean = str_remove(host_clean, "Golden Promise"),
    host_clean = str_remove(host_clean, "\\(pigeon pea\\)"),
    host_clean = str_remove(host_clean, "\\s+[A-Z][0-9]+$"),
    host_clean = str_trim(host_clean),
    host_clean = str_replace(host_clean, "^Glicyne", "Glycine"),
    host_clean = str_replace(host_clean, "^Desmonium", "Desmodium"),
    host_clean = str_replace(host_clean, "^Glycine hispida$", "Glycine max"),
    host_clean = if_else(
      host_clean %in% c("not applicable", "missing", "-", "None", "not provided",
                         "not recorded", "Not determined", "Plant", "legume",
                         "forest legume species") | is.na(host_clean),
      NA_character_, host_clean
    )
  )

genomes_curated <- genomes_curated |>
  mutate(
    host_genus = case_when(
      is.na(host_clean) ~ NA_character_,
      host_clean == "Homo sapiens" ~ "ANOMALY_review",
      host_clean == "Sus scrofa" ~ "ANOMALY_review",
      host_clean == "Subsurface shale" ~ "Non_plant_source",
      str_detect(host_clean, ";") ~ "MULTIPLE_review",
      str_detect(host_clean, regex("^soybean", ignore_case = TRUE)) ~ "Glycine",
      str_detect(host_clean, regex("^(forage )?peanut", ignore_case = TRUE)) ~ "Arachis",
      str_detect(host_clean, regex("pigeon\\s*pea", ignore_case = TRUE)) ~ "Cajanus",
      str_detect(host_clean, regex("^rice$", ignore_case = TRUE)) ~ "Oryza",
      str_detect(host_clean, regex("sweet potato", ignore_case = TRUE)) ~ "Ipomoea",
      str_detect(host_clean, regex("sugarcane", ignore_case = TRUE)) ~ "Saccharum",
      TRUE ~ word(host_clean, 1)
    ),
    host_genus_final = if_else(
      host_genus %in% c("ANOMALY_review", "MULTIPLE_review", "Non_plant_source"),
      NA_character_, host_genus
    )
  )

genomes_curated |> count(host_genus_final, sort = TRUE) |> print(n = 70)
```

**Cases excluded from the host variable** (kept in the overall dataset, see
`README_curadoria.md` for detailed justification): `Homo sapiens` (n=2),
`Sus scrofa` (n=1), `Subsurface shale` (n=1), multiple-host field (n=1).

---

# Step 5 — Selection funnel summary

```{r}
selection_funnel <- tibble(
  step = c(
    "Total genomes in NCBI (raw)",
    "After removing GCF/GCA duplicates",
    "After quality filter (completeness >=90%, contamination <5%)",
    "With country identified",
    "With continent assigned",
    "With host genus identified"
  ),
  n_genomes = c(
    nrow(genomes_raw),
    nrow(genomes_dedup),
    nrow(genomes_curated),
    sum(!is.na(genomes_curated$country_clean)),
    sum(!is.na(genomes_curated$continent) & genomes_curated$continent != "Not applicable (environmental source)"),
    sum(!is.na(genomes_curated$host_genus_final))
  )
)

selection_funnel
```

---

# Step 6 — Exporting results from this phase

```{r, eval=FALSE}
write_xlsx(genomes_curated, "bradyrhizobium_metadados_curados.xlsx")
write_xlsx(selection_funnel, "selection_funnel.xlsx")

genomes_curated |>
  pull(assembly_accession) |>
  write_lines("accessions_finais.txt")
```

---

# Step 7 — Disk size estimate (before download)

```{bash, eval=FALSE}
dataformat tsv genome --inputfile bradyrhizobium_genomes.jsonl \
  --fields accession,assmstats-total-sequence-len \
  > tamanhos_genomas.tsv
```

```{r, eval=FALSE}
genome_sizes <- read_tsv("tamanhos_genomas.tsv", show_col_types = FALSE) |>
  clean_names()

selected_sizes <- genomes_curated |>
  select(assembly_accession) |>
  left_join(genome_sizes, by = "assembly_accession")

total_gb_nucleotides <- sum(
  selected_sizes$assembly_stats_total_sequence_length, na.rm = TRUE
) / 1e9

total_gb_nucleotides  # ~14.74 GB of nucleotides
```

Estimate: ~14.74 GB of nucleotides + ~30% additional for proteins (~19 GB
uncompressed); downloaded `.zip` package expected around 5-7 GB.

---

# Step 8 — Downloading selected genomes

```{bash, eval=FALSE}
tmux new -s download_bradyrhizobium
conda activate bradyrhizobium
cd ~/bradyrhizobium_project

datasets download genome accession --inputfile accessions_finais.txt \
  --include genome,protein,seq-report \
  --filename bradyrhizobium_genomes.zip

# Ctrl+b, then d, to detach and leave it running in the background
```

```{bash, eval=FALSE}
unzip bradyrhizobium_genomes.zip -d bradyrhizobium_genomes/

find bradyrhizobium_genomes/ncbi_dataset/data/ -name "*.fna" | wc -l
```

**Status: complete.** 1727/1727 genomes downloaded and validated (100%
integrity check, `valid data package`). Two failed attempts occurred first
(HTTP/2 stream error; corrupted zip archive) — both resolved by simply
retrying the download.

---

# Step 9 — Fine-grained taxonomic classification (ANI against type genomes)

## Reference genomes

**Methodological note:** the three target groups are *B. japonicum*,
*B. elkanii*, and *B. iriomotense* — *B. diazoefficiens* is treated as part of
the *japonicum* group and therefore not used as a separate reference.
Accessions below were cross-validated directly against live NCBI data (not
assumed from external sources): all assemblies deposited under each species
name were queried, and candidates matching the type-strain identifier were
compared on assembly level, completeness and contamination. For *japonicum*
and *elkanii*, two independent sequencing projects exist for the same type
strain; the Complete-Genome-level assembly was selected over the
Scaffold-level alternative in both cases.

```{bash, eval=FALSE}
# Cross-validation query used for each species, e.g.:
datasets summary genome taxon "Bradyrhizobium elkanii" --as-json-lines | \
  dataformat tsv genome --fields accession,organism-name,assminfo-biosample-strain,assmstats-total-sequence-len

# Direct quality comparison between candidate accessions for the same strain:
datasets summary genome accession GCF_023278185.1,GCF_000379145.1 --as-json-lines | \
  dataformat tsv genome --fields accession,assminfo-level,checkm-completeness,checkm-contamination,assminfo-release-date
```

| Species | Type strain | Accession | Level | Completeness | Contamination |
|---|---|---|---|---|---|
| *Bradyrhizobium japonicum* | USDA 6 | GCF_000284375.1 | Complete Genome | 99.24% | 0.15% |
| *Bradyrhizobium elkanii* | USDA 76 | GCF_023278185.1 | Complete Genome | 99.4% | 0.42% |
| *Bradyrhizobium iriomotense* | EK05 (NBRC 102520) | GCF_030160715.1 | Contig | 99.61% | 1.2% |

```{bash, eval=FALSE}
datasets download genome accession GCF_000284375.1 --include genome --filename japonicum_ref.zip
datasets download genome accession GCF_023278185.1 --include genome --filename elkanii_ref.zip
datasets download genome accession GCF_030160715.1 --include genome --filename iriomotense_ref.zip

unzip japonicum_ref.zip -d ref_japonicum/
unzip elkanii_ref.zip -d ref_elkanii/
unzip iriomotense_ref.zip -d ref_iriomotense/

find bradyrhizobium_genomes/ncbi_dataset/data/ -name "*.fna" > lista_query_genomas.txt
find ref_japonicum/ ref_elkanii/ ref_iriomotense/ -name "*.fna" > lista_referencias.txt
```

## Running FastANI

**Environment note:** FastANI required a dedicated conda environment
(`fastani_env`), isolated from `bradyrhizobium`, due to a GSL (GNU Scientific
Library) shared-library version conflict between packages.

```{bash, eval=FALSE}
mamba create -n fastani_env -c bioconda -c conda-forge fastani -y
conda activate fastani_env

fastANI --ql lista_query_genomas.txt --rl lista_referencias.txt \
  -o resultado_ani_bradyrhizobium_v2.tsv --threads 4
```

**Status:** re-run in progress with the corrected reference set
(japonicum/elkanii/iriomotense, no diazoefficiens). A first version of this
comparison (including diazoefficiens as a fourth reference, and an earlier,
lower-quality elkanii/japonicum accession pair) was run previously; results
below refer to the corrected `_v2` run.

## Why a 95% ANI threshold for species assignment

The 95-96% ANI threshold used to assign species identity is not an arbitrary
choice — it is grounded in comparative genomics literature relating whole-
genome Average Nucleotide Identity to the pre-genomic "gold standard" species
definition based on DNA-DNA hybridization (DDH):

- **Goris et al. (2007)**, *International Journal of Systematic and
  Evolutionary Microbiology*, showed that the classical 70% DDH species
  threshold corresponds empirically to ~95-96% ANI across whole genomes.
- **Richter & Rosselló-Móra (2009)**, *PNAS*, reinforced ANI as a practical,
  sequence-based proxy for the prokaryotic species concept.
- **Jain et al. (2018)**, *Nature Communications* — the paper describing and
  validating **FastANI itself** — recommends and validates the ~95% ANI cutoff
  as the species boundary using the FastANI algorithm specifically. This is
  the most directly applicable reference for this analysis, since we use the
  same tool the threshold was validated with.

The threshold reflects an empirical discontinuity in the distribution of ANI
values across large genome collections (few genome pairs fall between ~83%
and ~95% ANI), rather than an exact mathematical cutoff — different authors
use values ranging from 94% to 96%, which is a normal source of variation in
microbial taxonomy, not disagreement about the underlying concept.

```{r, eval=FALSE}
ani_results <- read_tsv("resultado_ani_bradyrhizobium_v2.tsv",
                         col_names = c("query", "reference", "ani",
                                       "mapped_fragments", "total_fragments"),
                         show_col_types = FALSE)

nrow(ani_results)  # expected: 5181 (1727 genomes x 3 references)
head(ani_results)
```

```{r, eval=FALSE}
ani_results <- ani_results |>
  mutate(
    reference_group = case_when(
      str_detect(reference, "japonicum") ~ "japonicum",
      str_detect(reference, "elkanii") ~ "elkanii",
      str_detect(reference, "iriomotense") ~ "iriomotense"
    )
  )

best_match <- ani_results |>
  group_by(query) |>
  slice_max(ani, n = 1) |>
  ungroup() |>
  mutate(
    taxonomic_group = if_else(ani >= 95, reference_group, "Other_Bradyrhizobium")
  )

best_match |> count(taxonomic_group, sort = TRUE)
```

## Gray-zone check (genomes close to the 95% boundary)

Since the exact threshold varies between 94% and 96% depending on the
reference cited, it is worth checking how many genomes would change
classification depending on where exactly the cutoff is drawn — this
quantifies how sensitive the final result is to this specific methodological
choice.

```{r, eval=FALSE}
gray_zone <- best_match |>
  filter(ani >= 94 & ani < 96)

nrow(gray_zone)

gray_zone |> count(reference_group, sort = TRUE)
```

If `nrow(gray_zone)` is small relative to the total dataset, the threshold
choice has limited impact on the overall classification and does not need
extensive sensitivity analysis. If it is large, consider reporting results
under both a 94% and a 96% cutoff as a robustness check.

---

# Next steps (to document in this same file)

- [x] Confirm complete and intact download of the 1727 genomes
- [ ] Re-run fine-grained taxonomic classification via ANI with corrected
      reference set (japonicum/elkanii/iriomotense; diazoefficiens removed)
- [ ] Core-gene / MLSA extraction for phylogeny construction
- [ ] Phylogenetic tree construction (PhyloPhlAn or Roary + IQ-TREE)
- [ ] Generation of the three annotation maps over the phylogeny (continent,
      host, taxonomic group) via `ggtree`
- [ ] Phylogenetic signal tests (Pagel's lambda) and complementary statistics

---

# References

- Bowers RM, Kyrpides NC, Stepanauskas R, et al. (2017). Minimum information
  about a single amplified genome (MISAG) and a metagenome-assembled genome
  (MIMAG) of bacteria and archaea. *Nature Biotechnology*, 35, 725-731.
- Goris J, Konstantinidis KT, Klappenbach JA, et al. (2007). DNA-DNA
  hybridization values and their relationship to whole-genome sequence
  similarities. *International Journal of Systematic and Evolutionary
  Microbiology*, 57, 81-91.
- Jain C, Rodriguez-R LM, Phillippy AM, Konstantinidis KT, Aluru S (2018).
  High throughput ANI analysis of 90K prokaryotic genomes reveals clear
  species boundaries. *Nature Communications*, 9, 5114.
- Richter M, Rosselló-Móra R (2009). Shifting the genomic gold standard for
  the prokaryotic species definition. *PNAS*, 106(45), 19126-19131.
- Sobol MS, Klos AS, Ané C, McMahon KD, Kaçar B (2026). Ecological
  constraints and evolutionary trade-offs shape nitrogen fixation across
  habitats. *ISME Communications*, 6(1), ycag007.
